\chapter{Introduction and Motivation}

In the rapidly evolving field of software development,
managing development environments and ensuring consistent deployment
has become increasingly challenging \cite{ContainerTechnology}.
This thesis explores functional package management and containerization,
two key concepts that address these challenges.

Functional package management creates reproducible
and consistent software environments through a declarative approach.
It ensures that software components
and their dependencies are precisely defined, reducing conflicts and errors.
This method is crucial for projects where reproducing the exact development environment is essential
for debugging, testing, and deployment \cite{malkaIncreasingTrustOpen2024}.

Containerization, on the other hand, packages applications
and their dependencies into portable containers.
This ensures that applications run consistently across various environments,
from development to production.
Containers provide isolated environments for applications,
simplifying deployment and enhancing scalability \cite{pahlContainerizationPaaSCloud2015}.

The relevance of this thesis lies in examining these foundational concepts,
which are vital in modern software development.
As organizations adopt
DevOps \cite{DevOps2024} practices
and CI/CD \cite{CICD2024} pipelines,
the need for reliable tools to manage environments and deployments grows \cite{ContainerTechnology}.
Understanding functional package management
and containerization helps developers and organizations make informed decisions about their tools and methodologies.

The motivation for this thesis comes from the increasing complexity of software projects
and the challenges of managing development environments.
Issues like "dependency hell" \cite{DependencyHell2024}
and environment discrepancies can slow development, introduce bugs, and complicate deployment.
This thesis aims to provide solutions to these problems
by exploring functional package management and containerization.

Reproducibility is a significant driver for this study.
In complex projects, reliably recreating the development environment is crucial.
Functional package management ensures reproducibility,
while containerization guarantees consistent and portable deployment,
maintaining stability across different stages of the software lifecycle \cite{malkaIncreasingTrustOpen2024}.

This thesis offers a comprehensive understanding of functional package management
and containerization, filling a gap in existing literature.
The insights gained will help developers, system administrators,
and organizations choose the best tools for their needs.
As the software development landscape evolves,
understanding these concepts will be essential for managing environments and deployments effectively.

In conclusion, this thesis addresses the challenges of modern software development
by exploring functional package management and containerization.
By enhancing the efficiency, reliability, and scalability of development and deployment processes,
this work contributes valuable insights to both academic and industry audiences.

\input{src/chapters/01_Introduction/01_Objectives}
\input{src/chapters/01_Introduction/02_Approach}
\input{src/chapters/01_Introduction/03_Scope}
